---
title: "Prediction Assignment Writeup"
author: "SJ Oosthuizen"
date: "26/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Data description

The outcome variable is `classe`, a factor variable with 5 levels. For this data set, participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions:

- exactly according to the specification (Class A)
- throwing the elbows to the front (Class B)
- lifting the dumbbell only halfway (Class C)
- lowering the dumbbell only halfway (Class D)
- throwing the hips to the front (Class E)

## Initial configuration

The initial configuration consists of loading some required packages and initializing some variables.

```{r configuration, echo=TRUE, results='hide'}
#Data variables
training.file   <- './data/pml-training.csv'
training.url    <- 'https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv'
test.file <- './data/pml-testing.csv'
test.url  <- 'https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv'

#R Libraries
library(dplyr) #transforming data
library(ggplot2) #plotting
library(ggcorrplot) #plotting correlation matrix
library(caret) #machine learning
library(rattle) #plotting trees

# Seed for reproducability
set.seed(99)
```
## Data processing
In this section the data is downloaded and processed. Some basic transformations and cleanup will be performed, so that `NA` values are omitted. Removal of irrelevant columns such as `user_name`, `raw_timestamp_part_1`, `raw_timestamp_part_2`, `cvtd_timestamp`, `new_window`, and  `num_window` (columns 1 to 7)

The `pml-training.csv` data is used to devise training and testing sets.
The `pml-test.csv` data is used to predict and answer the 20 questions based on the trained model.

```{r dataprocessing, echo=TRUE, results='hide'}
# Download data
download.file(training.url, training.file)
download.file(test.url,test.file )
# Clean data
training   <-na.omit(read.csv(training.file))
testing <-na.omit(read.csv(test.file))
# Subset data
training   <-training[,-c(1:7)]
testing <-testing[,-c(1:7)]
```
## Cross-validation
In this section cross-validation will be performed by splitting the training data in training (75%) and testing (25%) data.

```{r datasplitting, echo=TRUE, results='hide'}
subSamples <- createDataPartition(y=training$classe , p=0.75, list=FALSE)
subTraining <- training[-subSamples,] 
subTesting <- training[-subSamples,]
```
## Exploratory analysis
The variable `classe` contains 5 levels. The plot of the outcome variable shows the frequency of each levels in the subTraining data.

```{r exploranalysis, echo=TRUE}
plot(subTraining$classe, subTraining$classe , col="red", main = "Levels of classes", xlab = "classe levels", ylab = "Frequency")
```

The plot above shows that Level A is the most frequent classe. D appears to be the least frequent one.


### Decision tree
```{r decisiontree, echo=TRUE}
# Fit model
modFitDT <- rpart(classe ~ ., data=subTraining, method="class")
# Perform prediction
predictDT <- predict(modFitDT, subTesting, type = "class")
# Plot result
rpart.plot(modFitDT, main="Classification Tree", extra=102, under=TRUE, faclen=0)
```

Following confusion matrix shows the errors of the prediction algorithm.

```{r decisiontreecm, echo=TRUE}
confusionMatrix(predictDT, subTesting$classe)
```
### Random forest
```{r randomforest, echo=TRUE}
# Fit model
modFitRF <- randomForest(classe ~ ., data=subTraining, method="class")
# Perform prediction
predictRF <- predict(modFitRF, subTesting, type = "class")
```

Following confusion matrix shows the errors of the prediction algorithm.

```{r randomforestcm, echo=TRUE}
confusionMatrix(predictRF, subTesting$classe)
```
## Submission
In this section the files for the project submission are generated using the random forest algorithm on the testing data.

```{r submission, echo=TRUE}
# Perform prediction
predictSubmission <- predict(modFitRF, testing, type="class")
predictSubmission
# Write files for submission
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("./data/submission/problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(predictSubmission)
